Gradual Soundness: Lessons from Static Python

### 2022-01-14 pj-7.1/paper.tex
pj: related work, no cites

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~[CITE] are two notable sources.
In addition, \SP{} is strongly influenced by the following engineering constraints:
\begin{itemize}
  \item
    Performance is the bottom line.
    At the end of the day, \SP{} needs to make code run faster in production.
  \item
    Critical code often relies on first-order functions and objects,
    at least in the Instagram web server.
    Precise types for higher-order functions and first-class classes
    are thus a low priority.
  \item
    The codebase in which \SP{} is being applied makes heavy use of Pyre [CITE]
    type annotations.
    \SP{} needs to be compatible with Pyre syntax to reduce the adoption burden.
    % On the flip side, the existence of Pyre lets \SP{} focus on a small language
    % of types that enable optimizations.
    % (\SP{} defers to Pyre the task of statically checking the full Python language.)
  \item
    A module-level granularity is acceptable.
    Once \SP{} is enabled for a module, it compiles all code in that module
    including expressions that have the dynamic type~(\cref{s:type-dynamic}).
    If this behavior is a problem for certain expressions, it is easy enough to move
    them to an untyped Python module.
\end{itemize}
%
In general, the types that \SP{} implements all enable significant optimizations
and can be validated with constant-time dynamic checks.
Types that do not meet these criteria are deferred to Pyre, even if they are widely-used.

@@

Concrete types

Nom~\cite{mt-oopsla-2017} and MonNom~\cite{mt-oopsla-2021} are foundational.
%% \footnote{In other words, \SP{} provides \emph{immediate accountability}~\cite{mt-oopsla-2017}.}

Full Monty~\cite{pmmwplck-oopsla-2013} core calculus for Python language, foundational for us.

Grace nominal transient aint bad at all

Thorn~\cite{wnlov-popl-2010} and StrongScript~\cite{rzv-ecoop-2015} are related.

PyPy is another runtime for Python
Pycket is built on PyPy and improves both Typed Racket~\cite{bbst-oopsla-2017}
and Reticulated Python~\cite{vsc-dls-2019}.

Type systems for Python ...
mypy pyre pytype;
Reticulated;

Optimizing type systems for Python ...
Reticulated;
mypyc;
Developer experience is one of the main advantages of \SP{} over mypyc,
which implements type-directed optimizations for CPython
by compiling source code to C extension modules.
\SP{} even used Cython at some point and found the mixed-workflow codebase painful.
%%  \item
%%    Prior to investing in \SP{}, Instagram used Cython to compile
%%    performance-critical code to C extension modules.
%%    Cython performed well, but mixing pure Python with a few compiled modules
%%    made for an awkward developer experience.
%%    \SP{} aims to match the Python workflow.

optimized transient, sp suggests method-at-a-time equally effective as tracing

JS++ is a sound gradual type system for JavaScript~\cite{jspp}.
By default, it separates JS++ from JavaScript values:
every value with a type must be defined in JS++ code,
and every value that may have come from Javascript has
a special type \code{external} to mark it as such.
Programmers can manually cast external values to a
more precise type.
The casts may fail at runtime, and may create a copy of
the original value if they succeed (depends on the type).
%% 2022-01-09 how does JS++ protect arrays sent to JS?

Gradual soundness in \SP{} bears a strong resemblance to \emph{progressive types} vision
of a type system that comes in strict and lax forms~\cite{pqk-onward-2012}.
Primitive types are progressive types (and a little more) for base values.
Concrete types are essentially a progressive form of shallow types.

\section{Future Work}
\label{s:future}

%% Immediate future: fix adoption friction points
%% 2021-12-15: WHAT ARE THESE?
%% - mocks need manual return_value ... "next half we plan to work on improving the Python mock framework to automatically create mocks of the correct types."
%% - mock calls have to rewrite to expect positional args, not keywords "plan to avoid rewrite if fn is patched"
%% - functions have to expect positionals (can't do keyword only)
%% - @dataclass enum.Enum
%%
%% significant selection bias
%% all of this is the low-hanging fruit, haven't tackled uses of metaclasses even though looks promising
%%  because the pay / payoff isn't yet attractive

%% engineering: JIT profiling, instead of hand-requested JIT list
%% model: metaclasses

%% 2021-12-30 Whence Nominal GT?
%%
%% KC: I think in general adding gradual typing to Java/C# (an also Python) is boring:
%% - MonNom allows you to cast between an arbitrary type and dyn.
%% - dyn is effectively Object
%% - Upcasts, casts from an arbitrary type to Object, are already in the language
%% - Downcasts, casts from Object to an arbitrary type is also possible, but must be explicit in the fully static language
%% So MonNom effectively gives programmers another way to name Object, with which downcasts can be automatically inserted by the compiler.
%% I think this sentence covers all the interesting stuff about MonNom.
%%
%% SK: This matches my intuition that gradual nominal is just not very
%% interesting. KC puts this on a technical footing with his very astute comment,
%% which to me can be summarized as "every nominal language that has Object is
%% already gradual" [leaving aside the boring non-Object base type distinction].
       So the entire remaining game is one of making the programming interface
%% as painless as possible. Some of that is just a trivial matter of
%% syntactic defaults; some of it is making the type-checker more forgiving
%% and replacing static checks with dynamic checks. There doesn't seem to
%% be much else left.

Adapt SP ideas to a new setting, test performance takeaways.

Improve SP with generic types, structural types (lambda), \ldots.

Use confined GT to relax ChkDict ... let programmers decide whether
the type should reject or convert untyped data.
Maybe a ChkDict function could compile to an "overloaded" version with
fast and slow paths for ChkDict vs normal dict.

Build an automatic migration tool for Checked data.

Build a static analysis that hoists transient annotations to an early, shared point.
Take care to give quality error messages.

Optimize Python integer operations (in addition to the c types).
Hard because syntax like \code{a + b} may be the result of either \code{a.\_\_add\_\_(b)}
or \code{b.\_\_radd\_\_(a)} depending on runtime types and behavior.

@@

@@

\subsection{Nominal, Checked Types}
\label{s:concrete-migration}

CheckedDict enables strong type checks and optimizations, but is painful to use.
It is painful even though ChkDict supports the full dict API
(any context that uses a Python dict can use a ChkDict without code changes).

The problem is that the ChkDict type rejects all Python values.
It accepts only ChkDict values, built through a special constructor
that installs a tag for type tests and guards writes.
Suppose that \code{f} is an untyped dict function.
As is, it can process ChkDict but gets no benefit from type checks nor from optimizations.
Adding a ChkDict domain type enables optimizations within \code{f} but raises a non-local
problem: all callers of \code{f} must be sure to create a ChkDict.
These ChkDicts must also be monomorphic to match whatever type \code{f} chose.
It can take many edits to get a program running again after adding a ChkDict type.

Lessons:
\begin{itemize}
  \item
    Nominal types are not compatible with structual values, such as Python values.
    Programmers are forced to edit old code to use such types ... unless the recent Nom work has a better idea~\cite{mt-oopsla-2021}
  \item
    Monomorphic nominal types are even worse.
    Editing old code to fit their rigid constraints may not be feasible.

Enabling checked types has a huge ripple effect.
The worst case is changing an annotation to a expect a checked data structure:
\begin{enumerate}
  \item all callers must create a checked value by invoking the right constructor, and
  \item all clients of those callers can no longer expect an unchecked type.
\end{enumerate}
Programmers have to trace the annotation back to value-creation points, and then
ensure that all uses of those values are compatible.

\subsection{Pyre Compatibility}

Many decisions in \SP{} are influenced by the widespread use of Pyre at Instagram.
\begin{itemize}
  \item
    To minimize changes needed to adapt Pyre-annotated programs
    to \SP{}, the syntax of \SP{} is compatible with Pyre.
    Pyre can run useful checks on every \SP{} program.
  \item
    Because Pyre is a mature type checker, the \SP{} team has taken a
    depth-first approach to development.
    The language implements a small set of sound types efficiently and ignores
    other types.
    Of course, programmers who use advanced Pyre types can still use Pyre to
    get compile-time feedback.
  \item
    The fact that \SP{} interprets unknown Pyre types as type dynamic changes
    users' expectations about what the dynamic type should mean.
    It is not necessarily a label for untyped Python code; dynamic code
    can be analyzed to catch static bugs.
    %% FILL what are the checks?
\end{itemize}

\subsection{Conclusion}

@@

\subsection{Communication}

How did we communicate effectively?
Starting point was GT survey [CITE].
Moved forward with example programs, see emails and github issues.
Our formalism had to cover their unit tests.
FILL

@@

FILL if we're going to list issues, might as well report the titles, labels, and open/closed status.

@@

>>>>

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Nom~\cite{mt-oopsla-2017,mt-oopsla-2021} and PEP 484~[CITE] are two notable sources.
In addition, \SP{} is strongly influenced by the following engineering constraints:
\begin{itemize}
  \item
    Performance is the bottom line.
    At the end of the day, \SP{} needs to make code run faster in production.
  \item
    Critical code often relies on first-order functions and objects,
    at least in the Instagram web server.
    Precise types for higher-order functions and first-class classes
    are thus a low priority.
  \item
    The codebase in which \SP{} is being applied makes heavy use of Pyre [CITE]
    type annotations.
    \SP{} needs to be compatible with Pyre syntax to reduce the adoption burden.
    % On the flip side, the existence of Pyre lets \SP{} focus on a small language
    % of types that enable optimizations.
    % depth-first types.
    % (\SP{} defers to Pyre the task of statically checking the full Python language.)
  \item
    A module-level granularity is acceptable.
    Once \SP{} is enabled for a module, it compiles all code in that module
    including expressions that have the dynamic type~(\cref{s:type-dynamic}).
    If this behavior is a problem for certain expressions, it is easy enough to move
    them to an untyped Python module.
\end{itemize}
%
In general, the types that \SP{} implements all enable significant optimizations
and can be validated with constant-time dynamic checks.
Types that do not meet these criteria are deferred to Pyre, even if they are widely-used.

@@

Concrete types originated in Thorn~\cite{bloom,wnlov-popl-2010}
and StrongScript~\cite{rzv-ecoop-2015},
and play a central role in Nom~\cite{mt-oopsla-2017} and MonNom~\cite{mt-oopsla-2021}.
It was thanks to Nom that \SP{} acquired concrete types; the impressive performance results
in the original Nom paper inspired the \SP{} team to pursue an implementation.

In addition to concrete types, Thorn and StrongScript have optional like types.
The mix of like and concrete types is similar to \SP{}'s mix of concrete and shallow
types, though like types make no claims about soundness.

Many aspects of our Redex model for \SP{} are adapted from the Full Monty core calculus
for Python~\cite{pmmwplck-oopsla-2013}.
In particular, the Full Monty paper reminded us that booleans \emph{are} integers in Python;
\SP{} had overlooked this detail.\footnote{Bug report: \url{https://github.com/facebookincubator/cinder/issues/46}}
%% 2022-01-14 TODO check full monty with KC

Our production experience with \SP{} suggests that a method-based JIT compiler can
usually eliminate the costs of shallow types.
This finding complements prior positive experience with shallow types and JIT
technology; e.g., in Reticulated with the pypy tracing JIT~\cite{vsc-dls-2019}
and in Grace with the Graal partial-evaluation-based JIT~\cite{FILL}.
Similarly, the Pycket team used a tracing JIT to reduce the costs of higher-order
checks~\cite{bbst-oopsla-2017}.
It remains to be seen whether a non-tracing JIT is effective for higher-order
casts.

\SP{} is one of many type systems for Python.
Other optional type checkers include Pyre, mypy, and PyType;
these systems do not ensure safe interactions with untyped Python.
Another sound type system is Reticulated,
which enforces structural types (rather than nominal) using its transient
semantics~\cite{vss-popl-2017,FILL}.
Other \emph{optimizing} type systems for Python
include Reticulated and mypyc.
The latter compiles Python source to C extension modules; unlike \SP{},
it does not offer a pure-Python developer experience.

Although sound gradual type systems are rare in industry, there are at
least three others besides \SP{}.
Dart 2 is a nominally-typed language with a dynamic type that is similar
to the top object type but allows implicit downcasts to every other type~\cite{FILL}.
C\# includes a dynamic type that defers type checking until runtime~\cite{bmt-ecoop-2010}.
JS++ is a sound type system for JavaScript that allows untyped values
using a catch-all \code{external} type~\cite{jspp}.
%% Programmers can manually cast external values to a more precise type.
%% The non-immediate casts make copies (how does that work for functions?!)
%% also, how does JS++ protect arrays sent to JS?

Gradual soundness in \SP{} bears a strong resemblance to \emph{progressive types} vision
of a type system that comes in strict and lax forms~\cite{pqk-onward-2012}.
Primitive types are progressive types for base values.
Concrete types are essentially a progressive form of shallow types.

%% \section{Future Work}
%% \label{s:future}
%% 
%% %% Immediate future: fix adoption friction points
%% %% - mocks need manual return_value ... "next half we plan to work on improving the Python mock framework to automatically create mocks of the correct types."
%% %% - mock calls have to rewrite to expect positional args, not keywords "plan to avoid rewrite if fn is patched"
%% %% - functions have to expect positionals (can't do keyword only)
%% %% - @dataclass enum.Enum
%% %%
%% %% significant selection bias toward low-hanging fruit
%% %% model: metaclasses
%% %% engineering: JIT profiling, instead of hand-requested JIT list
%% 
%% Adapt SP ideas to a new setting, test performance takeaways.
%% 
%% Improve SP with generic types, structural types (lambda), \ldots.
%% 
%% Use confined GT to relax ChkDict ... let programmers decide whether
%% the type should reject or convert untyped data.
%% Maybe a ChkDict function could compile to an "overloaded" version with
%% fast and slow paths for ChkDict vs normal dict.
%% 
%% Build an automatic migration tool for Checked data.
%% 
%% Build a static analysis that hoists transient annotations to an early, shared point.
%% Take care to give quality error messages.
%% 
%% Optimize Python integer operations (in addition to the c types).
%% Hard because syntax like \code{a + b} may be the result of either \code{a.\_\_add\_\_(b)}
%% or \code{b.\_\_radd\_\_(a)} depending on runtime types and behavior.

@@

%% 2021-12-30 Whence Nominal GT?
%%
%% KC: I think in general adding gradual typing to Java/C# (an also Python) is boring:
%% - MonNom allows you to cast between an arbitrary type and dyn.
%% - dyn is effectively Object
%% - Upcasts, casts from an arbitrary type to Object, are already in the language
%% - Downcasts, casts from Object to an arbitrary type is also possible, but must be explicit in the fully static language
%% So MonNom effectively gives programmers another way to name Object, with which downcasts can be automatically inserted by the compiler.
%% I think this sentence covers all the interesting stuff about MonNom.
%%
%% SK: This matches my intuition that gradual nominal is just not very
%% interesting. KC puts this on a technical footing with his very astute comment,
%% which to me can be summarized as "every nominal language that has Object is
%% already gradual" [leaving aside the boring non-Object base type distinction].
%%     So the entire remaining game is one of making the programming interface
%% as painless as possible. Some of that is just a trivial matter of
%% syntactic defaults; some of it is making the type-checker more forgiving
%% and replacing static checks with dynamic checks. There doesn't seem to
%% be much else left.

@@

@@

@@

@@

\section{Fine-Grained Benchmark Data}

FILL
